Comparing fluorometric methods ( in vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds

### 2024-04-15 manuscript/manuscript.qmd
continued edits

Field samples were collected from eight ponds in Rhode Island during October and November 2021 (@fig-map). Surface samples were collected in triplicate by wading into each pond to a minimum depth of two feet to avoid collecting near the sediment. If a surface scum was present, the scum was gently brushed aside to avoid being collected in the sample bottles. Samples were collected in acid-washed 1 L amber bottles, placed in a cooler with ice, and stored at 4°C until analysis. At the same location as sample collection, measurements of chlorophyll and phycocyanin were made using the AlgaeTorch and the FluoroSense.

@@

Water samples for phytoplankton identification and enumeration were decanted from field collected samples into 250 ml amber HDPE bottles and were preserved with 25% glutaraldehyde. Samples were stored at 4°C and shipped the week after collection on ice to Phycotech, Inc. A minimum of 400 natural units were counted per sample, and quantitative biovolume estimates were also performed for each sample. A full description of methods and a quality assurance plan are available from Phycotech, Inc. (https://www.phycotech.com/Portals/0/PDFs/PhycoTech_INFO_PACKET.pdf).

All phytoplankton samples were collected on 2021-10-06. Samples from Barber, Indian and Yawgoo were collected on 2021-10-06 and samples from Curran, Mashapaug, and Warwick were collected on 2021-10-22; fluorometry was performed with 24 hours of sample collection.

@@

To compare extracted concentrations and cell counts to the various instruments we used simple linear regression and scatterplots. For these regressions and scatterplots, the lab replicates were averaged, resulting in three field replicates for each of the waterbodies. We derived the coefficient of determination from the regressions as a measure of fit and also qualitatively compare the device measurements to the extracted concentrations and phytoplankton counts with scatterplots. We do not compare slopes nor look for a slope of 1 as we have little expectation that the extracted concentrations would equal the values from the various devices given the difference in units and the different RFU returns across the devices.

@@

Extracted chlorophyll measurements ranged from 3.2 ug/L at Indian Lake to 81.3 ug/L in Mashapaug Pond (@tbl-chlsummary). Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values (@fig-chlascatter).  The Algaetorch and Phycoprobe measurements exhibited a strong relationship with extracted chlorophyll concentrations with R^2^ values of 0.82 and 0.81, respectively.  The Cyanofluor (R^2^=0.35) and Trilogy in vivo (R^2^=0.34) measurements increased as extracted chlorophyll concentrations increased, but the relationship was not as strong compared to the other instruments.

@@

Extracted phycocyanin measurements ranged from 1.3 ug/L at Indian Lake to 75.0 ug/L in Mashapaug Pond (@tbl-phycosummary). The Algaetorch (R^2^=0.86) and Phycoprobe (R^2^=0.84) were strongly related to extracted phycocyanin concentrations (@fig-phycoscatter). Phycocyanin measurements for the Fluorosense (R^2^=0.96) and the Cyanofluor (R^2^=0.78) were also strongly related with extracted phycocyanin.

@@

Extracted chlorophyll *a* was the most strongly correlated with cyanobacterial cell counts (R^2^=0.92) followed by the algaetorch (R^2^=0.78) and the Phycoprobe (R^2^=0.59). The Cyanofluor (R^2^=0.12) and Trilogy in vivo module (0.11) were weakly correlated (@fig-chlcount). Phycocyanin measurements from all instruments were strongly correlated with cell counts (R^2^ between 0.9 and 0.97)(@fig-phycocount).

@@

We are grateful to EPA Reviewer 1, EPA Reviewer 2, EPA Reviewer 3, ACESD QAM, ACESD Supervisor, and Tim Gleason for constructive reviews of this paper. The views expressed in this article are those of the authors and do not necessarily represent the views or policies of the U.S. Environmental Protection Agency. Any mention of trade names, products, or services does not imply an endorsement by the U.S. Government or the U.S. Environmental Protection Agency. The EPA does not endorse any commercial products, services, or enterprises. This contribution is identified by the tracking number ORD-XXXXXX of the Atlantic Coastal Environmental Sciences Division, Office of Research and Development, Center for Environmental Measurement and Modeling, US Environmental Protection Agency. Lastly, special thanks to Gracie, Anastasia, and Lilac for cheer leading and support during challenging analysis times.

>>>>

Field samples were collected from six ponds in Rhode Island. (@fig-map). Samples from Barber, Indian and Yawgoo were collected on 2021-10-06 and samples from Curran, Mashapaug, and Warwick were collected on 2021-10-22. Surface samples were collected in triplicate by wading into each pond to a minimum depth of two feet to avoid collecting near the sediment. If a surface scum was present, the scum was gently brushed aside to avoid being collected in the sample bottles. Samples were collected in acid-washed 1 L amber bottles, placed in a cooler with ice, and stored at 4°C until analysis. At the same location as sample collection, measurements of chlorophyll and phycocyanin were made using the AlgaeTorch and the FluoroSense.

@@

All phytoplankton samples were collected on 2021-10-06. Water samples for phytoplankton identification and enumeration were decanted from field collected samples into 250 ml amber HDPE bottles and were preserved with 25% glutaraldehyde. Samples were stored at 4°C and shipped the week after collection on ice to Phycotech, Inc. A minimum of 400 natural units were counted per sample, and quantitative biovolume estimates were also performed for each sample. A full description of methods and a quality assurance plan are available from Phycotech, Inc. (https://www.phycotech.com/Portals/0/PDFs/PhycoTech_INFO_PACKET.pdf).

@@

To compare extracted concentrations and cell counts to the various instruments we used simple linear regression and scatterplots. For these regressions and scatterplots, the lab replicates were averaged, resulting in three field replicates for each of the waterbodies. We derived the coefficient of determination from the regressions as a measure of fit and also qualitatively compared the device measurements to the extracted concentrations and phytoplankton counts with scatterplots. We do not compare slopes nor look for a slope of 1 as we have little expectation that the extracted concentrations would equal the values from the various devices given the difference in units and the different RFU returns across the devices.

@@

Extracted chlorophyll measurements ranged from a mean of 1.3 ug/L in Indian Lake to a mean of 32.2 ug/L in Mashapaug Pond (@tbl-chlsummary). Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values (@fig-chlascatter).  The Algaetorch and Phycoprobe measurements exhibited a strong relationship with extracted chlorophyll concentrations with R^2^ values of 0.82 and 0.81, respectively.  The Cyanofluor (R^2^=0.35) and Trilogy in vivo (R^2^=0.34) measurements increased as extracted chlorophyll concentrations increased, but the relationship was not as strong compared to the other instruments.

@@

Extracted phycocyanin measurements ranged from a mean of 0.2 ug/L in Indian Lake and Barber Pond to 10.5 ug/L in Mashapaug Pond (@tbl-phycosummary). The Algaetorch (R^2^=0.86) and Phycoprobe (R^2^=0.84) were strongly related to extracted phycocyanin concentrations (@fig-phycoscatter). Phycocyanin measurements for the Fluorosense (R^2^=0.96) and the Cyanofluor (R^2^=0.78) were also strongly related with extracted phycocyanin.

@@

Extracted chlorophyll *a* was the most strongly correlated with cyanobacterial cell counts (R^2^=0.92) followed by the Algaetorch (R^2^=0.78) and the Phycoprobe (R^2^=0.59). The Cyanofluor (R^2^=0.12) and Trilogy in vivo module (0.11) were weakly correlated (@fig-chlcount). Phycocyanin measurements from all instruments were strongly correlated with cell counts (R^2^ between 0.9 and 0.97)(@fig-phycocount).

@@

We are grateful to Hilary Snook, Darryl Keith, Steve Rego, ACESD QAM, ACESD Supervisor, and Tim Gleason for constructive reviews of this paper. The views expressed in this article are those of the authors and do not necessarily represent the views or policies of the U.S. Environmental Protection Agency. Any mention of trade names, products, or services does not imply an endorsement by the U.S. Government or the U.S. Environmental Protection Agency. The EPA does not endorse any commercial products, services, or enterprises. This contribution is identified by the tracking number ORD-XXXXXX of the Atlantic Coastal Environmental Sciences Division, Office of Research and Development, Center for Environmental Measurement and Modeling, US Environmental Protection Agency. Lastly, special thanks to Gracie, Anastasia, and Lilac for cheer leading and support during challenging analysis times.
